---
title: "ext_birds"
author: Rob Cooke (roboke@ceh.ac.uk)
date: 13/10/2023
output: 
  html_notebook:
    toc: TRUE
editor_options: 
  chunk_output_type: inline
---

## Setup

```{r}

# see renv for all packages

library(dplyr)
library(raster)

# # save snapshot of packages to renv
# renv::snapshot()

# set plotting theme
ggplot2::theme_set(cowplot::theme_cowplot())

# helper functions

# function for probability-based rounding
prob_ro <- function(ro) {
  ifelse(floor(ro) + runif(n(), 0, 1) < ro, floor(ro) + 1, floor(ro))
}

# 10^x function from: https://stackoverflow.com/questions/11610377/how-do-i-change-the-formatting-of-numbers-on-an-axis-with-ggplot
fancy_scientific <- function(l) {
  # turn in to character string in scientific notation
  l <- format(l, scientific = TRUE)
  # leave zero as 0
  l <- gsub("0e\\+00","0",l)
  # quote the part before the exponent to keep all the digits
  l <- gsub("^(.*)e", "'\\1'e", l)
  # turn the 'e+' into plotmath format
  l <- gsub("e", "%*%10^", l)
  # return this as an expression
  parse(text = l)
}

```

## Raw data

Island data:<br/>
Weigelt et al., 2013 - https://www.pnas.org/content/110/38/15307

Extinct birds data:<br/>
Sayol et al., 2020 - https://www.science.org/doi/full/10.1126/sciadv.abb6095

Bird distributions:<br/>
Bird species distribution maps of the world. Version 2018.1. - http://datazone.birdlife.org/species/requestdis

Native rodents:<br/>
PHYLACINE 1.2: The Phylogenetic Atlas of Mammal Macroecology - https://doi.org/10.5281/zenodo.1250504

Gridded climatic data:<br/>
WorldClim; Hijmans et al., 2005 - https://www.worldclim.org/data/worldclim21.html<br/>
Elevation; Farr, 2007 - https://srtm.csi.cgiar.org/srtmdata/

Landmasses and archipelagos:<br/>
GADM version 3.6 - https://gadm.org/data.html

Human arrival dates:<br/>
see references in 'colz'

Extinction probabilities for possibly extinct species: 
Butchart et al., 2018 - https://doi.org/10.1016/j.biocon.2018.08.014

## Load in preprocessed data

Preprocessing code available upon request (roboke@ceh.ac.uk)

```{r}

# islands within archipelagos
isl_orig <- readr::read_csv("data/isl_arch.csv")
# ID_weigelt = ID number from Weigelt et al., 2013; Region = Region number (see Supplementary Fig. 3); Long = Longitude; Lat = Latitude; Archip_weigelt = Archipelago name from Weigelt et al., 2013; Archip = Archipelago/region name; Archip_1:Archip_7 = Alternative archipelago names; Island = Island name; Island_1 = Alternative island name; Area = Island area

# recorded fossil predictor data
pred_orig <- readr::read_csv("data/pred.csv")
# Archip = Archipelago name; res_eff = Research effort; Dist = Isolation distance; SLMP = Surrounding landmass; Elev = Elevation; Temp = Temperature; Prec = Precipitation; varT = Temperature variability; varP = Precipitation variability; SR_GAM = Archipelago plant richness; tot_area = Total area; sd_area = SD area; pa_rodents = Native rodents; hum = Human arrival; end_spp = Endemic bird species; ext_spp = Fossil bird extinctions; ant_spp = Bird species extant at 1500 CE (extant and observed extinctions [post-1500 CE])

# upper bound predictor data
dri_orig <- readr::read_csv("data/dri.csv")
# x = Longitude; y = Latitude; Elev = Elevation; Temp = Temperature; Prec = Precipitation; varT = Temperature variability; varP = Precipitation variability; all_spp = Number of bird species

# archipelago predictor data
dri_arch_orig <- readr::read_csv("data/dri_arch.csv")
# Archip = Archipelago name; Elev = Elevation; Temp = Temperature; Prec = Precipitation; varT = Temperature variability; varP = Precipitation variability

# MCMC samples from white noise linear model
# includes 20,000 samples (used 600,000 samples in study)
mcmc_lin_orig <- readr::read_delim("data/py_mcmc_samples_white_noise_multi_regression_linear_model_interaction_area_reseffort.txt")

# scaled predictors used in white noise linear model
scaled_pred_orig <- readr::read_csv("data/scaled_pred.csv")

# human arrival dates
colz_orig <- readr::read_csv("data/colz.csv")
# Archip = Archipelago/region name; prehist = Prehistorically settled (i.e., pre-1500 CE); cont = Continental; hum = Central human arrival date estimate (years before present); hum_low = Lower estimate of human arrival date (years before present); hum_upp = Upper estimate of human arrival date (years before present); ref = Reference; full_ref = Full reference 

# bird extinctions
bird_orig <- readr::read_csv("data/bird.csv")
# Archip = Archipelago name; sp_id = Species ID; IDsps = ID from Sayol et al.; species = Species name; commonName = Common name; grp = Group; ext_date_min = Minimum extinction date; ext_date_max = Maximum extinction date; ext_date = Central extinction date; year_last_recorded = Year last recorded for possibly extinct species; p_records = Extinction probability based on records for possibly extinct species; p_threats = Extinction probability based on threats for possibly extinct species; fossil_max = Maximum age of fossil; fossil_min = Minimum age of fossil

```

## Modelling

Transformations and standardization

```{r}

pred <- pred_orig %>% 
  # drop extant species
  dplyr::select(-ant_spp)

# transformations
pred_trans_unscaled <- pred %>% 
  dplyr::mutate_at(vars(Dist, tot_area, SR_GAM, hum, Elev), log) %>% 
  dplyr::mutate_at(vars(sd_area, res_eff, end_spp), ~log(. + 1)) 

# centered and scaled to zero mean and unit variance
pred_trans <- pred_trans_unscaled %>% 
  dplyr::mutate_at(vars(-Archip, -ext_spp), scale) %>% 
  dplyr::mutate_at(vars(-Archip, -ext_spp), as.numeric)

```

#### Linear model (LM)

```{r}

# log-transform extinct species
pred_trans_log <- dplyr::mutate(pred_trans, ext_spp = log(ext_spp + 1))

# model formula
fmla <- ext_spp ~ Dist + Elev + Temp + varT + Prec + varP + SR_GAM + tot_area + pa_rodents + end_spp + res_eff + hum + tot_area:res_eff

## Linear

# run model
m1 <- lm(fmla, data = pred_trans_log, na.action = na.pass)

# model summary
jtools::summ(m1)

# coefficients and partial r2
coef <- rsq::rsq.partial(m1, type = "sse") %>% 
  data.frame() %>% 
  dplyr::mutate(partial.rsq_round = round(partial.rsq, digits = 2)) %>% 
  dplyr::full_join(broom::tidy(m1, conf.int = TRUE), by = c("variable" = "term"))

```

#### GLM

```{r}

m2 <- glm(fmla, data = pred_trans, family = poisson)

sim_fmp <- DHARMa::simulateResiduals(m2)
DHARMa::testDispersion(sim_fmp)
# dispersion = 4.2, p = 0

m3 <- glm(fmla, data = pred_trans, family = quasipoisson)

```

###### Cross-validation of predictive accuracy

```{r}

# leave-one-out-cross-validation

# loocv function
loocv <- function(i, df, mod) {

  # train data
  train <- dplyr::filter(df, Archip != i)
  # test data
  test <- dplyr::filter(df, Archip == i)
  
  # train model
  up <- update(mod, data = train)
  # test model
  pd <- predict(up, newdata = test, type = "response")
  
}

# cross-validate m1
cv1 <- sapply(pred_trans_log$Archip, function(i) loocv(i = i, df = pred_trans_log, mod = m1)) %>% 
  # back-transform
  exp(.) - 1

# goodness-of-fit m1
met1 <- hydroGOF::gof(cv1, exp(pred_trans_log$ext_spp) - 1, do.spearman = TRUE)

# cross-validate m3
cv3 <- sapply(pred_trans$Archip, function(i) loocv(i = i, df = pred_trans, mod = m3))

# goodness-of-fit m3
met3 <- hydroGOF::gof(cv3, pred_trans$ext_spp, do.spearman = TRUE)

```

## Extrapolations

#### Extrapolate values LM

```{r}

# research effort in Aotearoa New Zealand
nz <- pred_trans %>%
  dplyr::filter(Archip == "Aotearoa New Zealand") %>%
  dplyr::select(res_eff)

# 10,000 posterior draws for example (60,000 used for study)
draw <- 10000

# simulate the posterior coefficients
post_coef <- arm::sim(m1, draw) %>%
  .@coef %>%
  as.data.frame()

extrapolate_samps <- function(x, df) {

  # archipelago
  ar <- pred_trans_log[x,]$Archip

  # data for archipelago
  nd <- pred_trans_log %>%
    dplyr::filter(Archip == ar)

  # fossil extinct
  pre_spp <- nd$ext_spp

  # delta research effort (compared to New Zealand)
  delta_res <- nz - nd$res_eff

  # for each posterior sample
  out <- lapply(1:draw, function(b) {

    # total prehistoric extinct
    tot_spp <- pre_spp + delta_res$res_eff * (post_coef[["res_eff"]][b] + post_coef[["tot_area:res_eff"]][b] * nd$tot_area)

    # transform to natural scale
    tot_spp <- exp(tot_spp) - 1

    # subtract fossil extinct
    ext_spp <- tot_spp - (exp(pre_spp) - 1)

  })

  # extract data and tidy
  out_df <- unlist(out) %>%
    as.data.frame() %>%
    setNames(ar)

  # # print archipelago name - progress
  # print(ar)

  # return data
  return(out_df)

}

# extrapolate for each archipelago for each posterior estimate
pred_dist <- lapply(1:nrow(pred_trans_log), extrapolate_samps)

pred_uni <- dplyr::bind_cols(pred_dist) %>%
  tidyr::gather(key = "Archip", value = "pred_spp") %>%
  # drop Aotearoa New Zealand
  dplyr::filter(Archip != "Aotearoa New Zealand") %>%
  dplyr::mutate(sample = rep(1:(nrow(.)/(nrow(pred) - 1)), times = (nrow(pred) - 1))) %>%
  # probability-based rounding
  dplyr::mutate(pred_spp_r = prob_ro(pred_spp)) %>%
  dplyr::mutate(pred_spp_r = ifelse(pred_spp_r < 0, 0, pred_spp_r))

```

#### Extrapolate values GLM

```{r}

# simulate the posterior coefficients
post_coef <- arm::sim(m3, draw) %>%
  .@coef %>%
  as.data.frame()

# extrapolate for each archipelago for each posterior estimate
pred_dist <- lapply(1:nrow(pred_trans_log), extrapolate_samps)

pred_uni_glm <- dplyr::bind_cols(pred_dist) %>%
  tidyr::gather(key = "Archip", value = "pred_spp") %>%
  # drop Aotearoa New Zealand
  dplyr::filter(Archip != "Aotearoa New Zealand") %>%
  dplyr::mutate(sample = rep(1:(nrow(.)/(nrow(pred) - 1)), times = (nrow(pred) - 1))) %>%
  # probability-based rounding
  dplyr::mutate(pred_spp_r = prob_ro(pred_spp)) %>%
  dplyr::mutate(pred_spp_r = ifelse(pred_spp_r < 0, 0, pred_spp_r))

```

## Upper bound

#### Upper bound model

```{r}

# transformations
dri <- dri_orig %>% 
  dplyr::mutate(Elev = log(Elev + 1))

# scales for predictions
mean_scale <- sapply(dplyr::select(dri, Elev, Temp, Prec, varT, varP), mean)
sd_scale <- sapply(dplyr::select(dri, Elev, Temp, Prec, varT, varP), sd)

# centered and scaled to zero mean and unit variance
dri_scale <- dri %>% 
  dplyr::mutate_at(vars(Elev, Temp, Prec, varT, varP), ~as.numeric(scale(.)))

# non-spatial model 1
um1 <- glm(all_spp ~ Elev + Temp + Prec + varT + varP, family = poisson, data = dri_scale)

sim_fmp <- DHARMa::simulateResiduals(um1)
DHARMa::testDispersion(sim_fmp)
# dispersion = 55, p = 0

# non-spatial model 2
um2 <- glm(all_spp ~ Elev + Temp + Prec + varT + varP, family = quasipoisson, data = dri_scale)

# RAC model (autocovariate derived from residuals of model with environmental predictors)

xy <- cbind(dri_scale$x, dri_scale$y)
xyz <- cbind(xy, rep(NA, nrow(xy)))

rast <- raster::rasterFromXYZ(xyz, res = c(113000, 113000), crs = sp::CRS("+proj=eck4 +lon_0=0 +x_0=0 +ellps=WGS84 +units=m +no_defs"))

xy_residuals <- cbind(xy, resid(um2))

rast[raster::cellFromXY(rast, xy_residuals)] <- xy_residuals[,3]

focal_rac_rast <- raster::focal(rast, matrix(1, nrow = 3, ncol = 3), fun = mean, na.rm = TRUE, pad = TRUE)

focal_rac_vect <- raster::extract(focal_rac_rast, xy)

dri_scale_spa <- cbind(dri_scale, focal_rac_vect)

# RAC model
um2_rac <- glm(all_spp ~ Elev + Temp + Prec + varT + varP + focal_rac_vect, family = quasipoisson, data = dri_scale_spa)

# model summary
jtools::summ(um2_rac)

# leave-one-out-cross-validation

# loocv function
loocv_upp <- function(i, df, mod) {

  # training data
  train <- dplyr::filter(df, rowid != i)
  # test data
  test <- dplyr::filter(df, rowid == i)
  
  # train model
  up <- update(mod, data = train)
  # test model
  pd <- predict(up, newdata = test, type = "response")
  
}

# cross-validate upper bound model - takes a while
cv_um2 <- sapply(1:nrow(dri_scale), function(i) loocv_upp(i = i, df =
                                                            dri_scale_spa %>%
                  tibble::rownames_to_column("rowid"), mod = um2_rac))

met_um2 <- hydroGOF::gof(cv_um2, dri_scale$all_spp, do.spearman = TRUE)

```

#### Upper bound predictions for focal archipelagos

```{r}

# tidy data
dri_arch <- dri_arch_orig %>% 
  dplyr::mutate(Elev = log(Elev + 1)) %>% 
  # add residuals autocovariate
  dplyr::mutate(focal_rac_vect = median(focal_rac_vect))

# scale data to match model data
dri_arch_scale <- scale(dplyr::select(dri_arch, -Archip, -focal_rac_vect), center = mean_scale, scale = sd_scale) %>% 
  as.data.frame() %>% 
  # rejoin unscaled columns
  dplyr::bind_cols(dplyr::select(dri_arch, Archip, focal_rac_vect), .) %>% 
  # dummy variable needed for model.matrix function
  dplyr::mutate(all_spp = 0)

# function to get posterior distribution of model parameters
# code adapted from https://rdrr.io/cran/ciTools/src/R/add_pi_glm.R
get_sim_response <- function(df, fit, nSims) {

    nPreds <- NROW(df)
    modmat <- model.matrix(fit, data = df)
    # model distribution
    response_distr <- fit$family$family
    # inverse model link
    inverselink <- fit$family$linkinv
    # overdispersion parameter
    overdisp <- summary(fit)$dispersion
    # draw the posterior coefficients
    sims <- arm::sim(fit, n.sims = nSims)
    # set up matrix
    sim_response <- matrix(NA, ncol = nSims, nrow = nPreds)

    for (i in 1:nSims){
        yhat <- inverselink(modmat %*% sims@coef[i,])
        a <- yhat / (overdisp - 1)
        sim_response[,i] <- MASS::rnegbin(n = nPreds,
                                    mu = yhat,
                                    theta = a)
    }
    
  return(sim_response)
    
}

# 10,000 posterior draws for example (150,000 used for study)
upp_draw <- 10000

# posterior samples
upp_sim <- get_sim_response(df = dri_arch_scale, fit = um2_rac, nSims = upp_draw)

upp_samp_full <- dplyr::bind_cols(Archip = dri_arch_scale$Archip, as.data.frame(upp_sim)) %>%
  dplyr::group_by(Archip) %>%
  dplyr::summarise_if(is.numeric, median) %>%
  # wide to long
  tidyr::gather(key = "sample", value = "upp", contains("V"))

# adjust for archipelago area
upp_samp_area <- upp_samp_full %>%
  # join archipelago area, fossil bird extinctions, and extant and observed extinct (i.e., extant at 1500 CE) species
  dplyr::left_join(dplyr::select(pred_orig, Archip, tot_area, ext_spp, ant_spp), by = "Archip") %>%
  # log transform area - cell area 12,769 km^2
  dplyr::mutate(gce = log(tot_area / 12769)) %>%
  # species-area relationship
  # log(S) = log(c) + zlog(A)
  dplyr::mutate(upp_area = exp(log(upp) + (0.25 * (gce)))) %>%
  dplyr::mutate(sample = as.integer(gsub("V", "", sample)))

upp_samp <- upp_samp_area %>% 
  # total recorded (fossil extinct, observed extinct, extant)
  dplyr::mutate(rec_spp = ext_spp + ant_spp) %>% 
  # potential additional species
  dplyr::mutate(add_spp = upp_area - rec_spp) %>% 
  # turn negative numbers into zeros
  dplyr::mutate(add_spp = ifelse(add_spp < 0, 0, add_spp))

```

## Rejection sampling LM

```{r}

# predictions = pred_uni
# upper bound = upp_samp

comb_df <- pred_uni %>%
  dplyr::left_join(upp_samp, by = c("Archip", "sample")) %>%
  # total including undiscovered
  dplyr::mutate(tot_spp = rec_spp + pred_spp_r)

# upper bound rejection sampling
comb_upp_rej <- comb_df %>%
  # rejection
  dplyr::filter(pred_spp_r <= add_spp)

# percentage rejected
upp_perc <- comb_upp_rej %>%
  dplyr::count(Archip) %>%
  dplyr::mutate(perc = 100 - ((n / (nrow(comb_df)/length(unique(comb_df$Archip)))) * 100))

# upper bound rejection rate
median(upp_perc$perc)
mean(upp_perc$perc)

# 100 samples to keep (1,000 samples used in study)
n_samp <- 100

# tidy rejection sampling
comb_ex <- comb_upp_rej %>%
  # 100 estimates to take forward (1,000 samples used in study)
  dplyr::group_by(Archip) %>%
  dplyr::sample_n(n_samp) %>%
  # add run identifier
  dplyr::mutate(run = 1:n_samp) %>%
  dplyr::select(Archip, run, pred_spp_r, ext_spp, ant_spp, tot_spp)

# add data for Aotearoa New Zealand
nz_run <- dplyr::filter(pred_orig, Archip == "Aotearoa New Zealand") %>% 
  dplyr::mutate(pred_spp_r = 0) %>% 
  dplyr::mutate(tot_spp = ext_spp + ant_spp + pred_spp_r) %>% 
  dplyr::slice(rep(1:n(), each = n_samp)) %>% 
  dplyr::mutate(run = 1:n_samp) %>% 
  dplyr::select(Archip, run, pred_spp_r, ext_spp, ant_spp, tot_spp)

ex <- dplyr::bind_rows(comb_ex, nz_run)

# # load: ex
# ex <- readRDS("data/ex.rds") # this is the full dataset from the study

```

## Alternative model of fossil extinct birds

#### Extrapolations for white noise model

```{r}

# extrapolate

scaled_pred <- scaled_pred_orig
mcmc_lin <- mcmc_lin_orig

# research effort in Aotearoa New Zealand
nz <- scaled_pred %>%
  dplyr::filter(Archip == "Aotearoa New Zealand") %>%
  dplyr::select(res_eff)

# posterior coefficients
post_coef_wn <- dplyr::filter(mcmc_lin, iteration > 10000000) # remove burn-in

extrapolate_samps_wn <- function(x) {

  # archipelago
  ar <- scaled_pred[x,]$Archip

  # data for archipelago
  nd <- scaled_pred %>%
    dplyr::filter(Archip == ar)

  # recorded prehistoric extinct
  pre_spp <- (nd$ext_spp - min(scaled_pred$ext_spp)) / diff(range(scaled_pred$ext_spp))

  # delta research effort (compared to New Zealand)
  delta_res <- nz - nd$res_eff

  # for each posterior sample
  out <- lapply(1:nrow(post_coef_wn), function(b) {

    # total prehistoric extinct
    tot_spp <- pre_spp + delta_res$res_eff * (post_coef_wn[["m16"]][b] + post_coef_wn[["m17"]][b] * nd$tot_area)

    # transform to natural scale
    tot_spp <- tot_spp * diff(range(scaled_pred$ext_spp)) + min(scaled_pred$ext_spp)

    # subtract recorded prehistoric extinct
    ext_spp <- tot_spp - nd$ext_spp

  })

  # extract data and tidy
  out_df <- unlist(out) %>%
    as.data.frame() %>%
    setNames(ar)

  # # print archipelago name - progress
  # print(ar)

  # return data
  return(out_df)

}

# extrapolate for each archipelago for each posterior estimate
pred_dist_wn <- lapply(1:nrow(scaled_pred), extrapolate_samps_wn)

pred_uni_wn <- dplyr::bind_cols(pred_dist_wn) %>%
  tidyr::gather(key = "Archip", value = "pred_spp") %>%
  # drop Aotearoa New Zealand
  dplyr::filter(Archip != "Aotearoa New Zealand") %>%
  dplyr::mutate(sample = rep(1:(nrow(.)/(nrow(scaled_pred) - 1)), times = (nrow(scaled_pred) - 1))) %>%
  # probability-based rounding
  dplyr::mutate(pred_spp_r = prob_ro(pred_spp)) %>%
  dplyr::mutate(pred_spp_r = ifelse(pred_spp_r < 0, 0, pred_spp_r))

```

#### Rejection sampling Gaussian white noise

```{r}

# predictions = pred_uni_wn
# upper bound = upp_samp

comb_df_wn <- pred_uni_wn %>%
  # filter to match number of samples to upper bound
  dplyr::filter(sample %in% c(1:(nrow(upp_samp)/length(unique(upp_samp$Archip))))) %>%
  dplyr::left_join(upp_samp, by = c("Archip", "sample")) %>%
  # total including undiscovered
  dplyr::mutate(tot_spp = rec_spp + pred_spp_r)

# upper bound rejection sampling
comb_upp_rej_wn <- comb_df_wn %>%
  # rejection
  dplyr::filter(pred_spp_r <= add_spp)

# percentage rejected
upp_perc_wn <- comb_upp_rej_wn %>%
  dplyr::count(Archip) %>%
  dplyr::mutate(perc = 100 - ((n / (nrow(comb_df_wn)/length(unique(comb_df_wn$Archip)))) * 100))

# upper bound rejection rate
median(upp_perc_wn$perc)
mean(upp_perc_wn$perc)

# tidy rejection sampling
comb_ex_wn <- comb_upp_rej_wn %>%
  # 100 estimates to take forward (1,000 samples used in study)
  dplyr::group_by(Archip) %>%
  dplyr::sample_n(n_samp) %>%
  # add run identifier
  dplyr::mutate(run = 1:n_samp) %>%
  dplyr::select(Archip, run, pred_spp_r, ext_spp, ant_spp, tot_spp)

ex_wn <- dplyr::bind_rows(comb_ex_wn, nz_run)

# # load: ex_wn
# ex_wn <- readRDS("data/ex_wn.rds") # this is the full dataset from the study

```

## Extinction chronology

```{r}

colz <- colz_orig
bird <- bird_orig

# add proportional uncertainty to human arrival dates
colz_uncert <- colz %>% 
  # only archipelagos
  dplyr::filter(cont == 0) %>% 
  # average proportional uncertainty
  dplyr::mutate(hum_range = hum_low - hum_upp) %>% 
  dplyr::mutate(hum_uncert = hum_range / hum) %>% 
  # lower estimate
  dplyr::mutate(hum_low = ifelse(is.na(hum_low), hum + (mean(.$hum_uncert, na.rm = TRUE) * hum), hum_low)) %>% 
  # upper estimate
  dplyr::mutate(hum_upp = ifelse(is.na(hum_upp), hum - (mean(.$hum_uncert, na.rm = TRUE) * hum), hum_upp))

# randomized extinction chronology based on Aotearoa New Zealand (i.e., island)

# rate - half-life of 100 years
rate <- -log(0.5)/100

# point to truncate exponential - 90% of extinctions within 332 years
trunc <- -log(1/10)/rate

# truncated exponential
# 1,000,000 random numbers (100,000,000 used in study)
exp_trunc <- rexp(1000000, rate = rate)
exp_trunc <- exp_trunc[exp_trunc < trunc]

# divide into chunks of length 100,000 (10,000,000 used in study)
m <- 100000
x <- seq_along(exp_trunc)
exp_trunc_split <- split(exp_trunc, ceiling(x / m))

# randomized extinction chronology based on North America (i.e., continental)

# continental rate - half-life of 1,000 years
rate_cont <- -log(0.5)/1000

# point to truncate exponential - 90% of extinctions within 3,322 years
trunc_cont <- -log(1/10)/rate_cont

# truncated exponential
# 1,000,000 random numbers (10,000,000 used in study)
exp_trunc_cont <- rexp(1000000, rate = rate_cont)
exp_trunc_cont <- exp_trunc_cont[exp_trunc_cont < trunc_cont]

# divide into chunks of length 100,000 (1,000,000 used in study)
m <- 100000
x <- seq_along(exp_trunc_cont)
exp_trunc_cont_split <- split(exp_trunc_cont, ceiling(x/m))

# truncated exponential - Madagascar only
exp_trunc_cont_mad <- exp_trunc_cont_split[[1]]
# 75th quantile, 2,000 years, maximum of 1950 for Madagascar
exp_trunc_cont_mad <- exp_trunc_cont_mad[exp_trunc_cont_mad < -log(2.5/10)/rate_cont]

# human settlement of archipelagos
ex_da_colz <- ex %>% 
  dplyr::left_join(colz_uncert, by = "Archip") %>%
  # human first arrival from uniform distribution of dates - archipelago level
  dplyr::mutate(hum_unif = purrr::pmap(list(x = hum_upp, y = hum_low), ~ runif(1, .x, .y))) %>% 
  tidyr::unnest(cols = c(hum_unif)) %>% 
  dplyr::select(Archip:tot_spp, hum_unif)

# undiscovered

# undiscovered
ex_da_un <- ex_da_colz %>% 
  # create row for every species, keep count column
  tidyr::uncount(pred_spp_r, .remove = FALSE) %>%
  # extinction dates from truncated exponential - species level
  # Madagascar with continental rate extinction chronology
  dplyr::mutate(pred_date = ifelse(!Archip == "Madagascar", hum_unif - exp_trunc_split[[1]][1:n()], hum_unif - exp_trunc_cont_mad[1:n()])) %>% 
  # round down to integer, i.e., year
  dplyr::mutate(pred_date = floor(pred_date)) %>% 
  # distribution
  dplyr::mutate(dist = "pre_un")

# fossil

# fossil extinct birds
ex_da_kno <- ex_da_colz %>% 
  # create row for every species, keep count column
  tidyr::uncount(ext_spp, .remove = FALSE) %>%
  # extinction dates from truncated exponential - species level
  dplyr::mutate(pred_date = hum_unif - exp_trunc_split[[2]][1:n()]) %>% 
  # round down to integer, i.e., year
  dplyr::mutate(pred_date = floor(pred_date)) %>% 
  # distribution
  dplyr::mutate(dist = "pre_kno") %>% 
  # add species names
  dplyr::group_by(Archip, run) %>% 
  dplyr::mutate(sp_id = 1:n()) %>% 
  # fossil extinctions on focal archipelagos
  dplyr::left_join(dplyr::filter(bird, grp == "fossil"), by = c("Archip", "sp_id")) %>% 
  dplyr::select(Archip:commonName)

# combine datasets
ex_da <- dplyr::bind_rows(ex_da_un, ex_da_kno)

# observed extinct species without date of the last reliable or confirmed record - 17 species

hist_ins_all <- bird %>% 
  # observed extinct species without date of the last reliable or confirmed record
  dplyr::filter(grp == "obs_ins") %>% 
  # number of extinct birds per archipelago
  dplyr::count(Archip) %>% 
  # join colonization
  dplyr::left_join(colz_uncert, by = "Archip") %>% 
  # replicate dataframe 100 times (1,000 used in study)
  dplyr::slice(rep(1:n(), each = n_samp)) %>% 
  # add run identifier
  dplyr::mutate(run = rep(1:n_samp, n()/n_samp)) %>% 
  # human first arrival from uniform distribution of dates
  dplyr::mutate(hum_unif = purrr::pmap(list(x = hum_upp, y = hum_low), ~ runif(1, .x, .y))) %>% 
  tidyr::unnest(cols = c(hum_unif))

hist_ins <- hist_ins_all %>% 
  # create row for every species
  tidyr::uncount(n) %>% 
  dplyr::mutate(pred_date = hum_unif - exp_trunc_split[[3]][1:n()]) %>% 
  # round down to integer, i.e., year
  dplyr::mutate(pred_date = floor(pred_date)) %>% 
  # distribution
  dplyr::mutate(dist = "hist_kno") %>% 
  # add species names
  dplyr::group_by(Archip, run) %>% 
  dplyr::mutate(sp_id = 1:n()) %>% 
  # observed insular extinctions
  dplyr::left_join(dplyr::filter(bird, grp == "obs_ins"), by = c("Archip", "sp_id")) %>% 
  dplyr::select(Archip, run, hum_unif, pred_date, dist, species, commonName)

# observed extinct birds with dates

dates_hist_ext_long <- bird %>% 
  # observed extinct birds with dates
  dplyr::filter(grp == "obs_date") %>% 
  # replicate dataframe
  dplyr::slice(rep(1:n(), each = n_samp)) %>% 
  # add run identifier
  dplyr::mutate(run = rep(1:n_samp, n()/n_samp)) %>% 
  # add uncertainty from uniform distribution
  dplyr::mutate(ext_date = purrr::pmap(list(x = ext_date_min, y = ext_date_max), ~ runif(1, .x, .y))) %>% 
  tidyr::unnest(cols = c(ext_date)) %>% 
  # convert to years before present
  dplyr::mutate(pred_date = floor((1950 - ext_date))) %>% 
  dplyr::mutate(dist = "hist_kno")

# observed possibly extinct - 46 species
hist_pex <- bird %>% 
  # possibly extinct
  dplyr::filter(grp == "pex") %>% 
  # extinction date in years BP
  dplyr::mutate(pred_date = 1950 - year_last_recorded) %>% 
  # average extinction probability
  dplyr::mutate(prob_ext = 1 - rowMeans(dplyr::select(., p_records, p_threats))) %>% 
  # distribution
  dplyr::mutate(dist = "hist_un")

hist_pex_long <- hist_pex %>% 
  # replicate dataframe
  dplyr::slice(rep(1:n(), each = n_samp)) %>% 
  # add run identifier
  dplyr::mutate(run = rep(1:n_samp, n()/n_samp)) %>% 
  # binomial probability of extinction
  dplyr::mutate(pex = purrr::pmap(list(x = prob_ext), ~ rbinom(1, 1, .x))) %>% 
  tidyr::unnest(cols = c(pex)) %>%
  # just those that go extinct
  dplyr::filter(pex == 1) %>% 
  dplyr::select(Archip, species, commonName, run, pred_date, dist)

# all but palearctic and indo-malay
cont_pre_excl_all <- bird %>% 
  # continental fossil extinctions
  dplyr::filter(grp == "cont") %>% 
  # excluding palearctic and indo-malay
  dplyr::filter(!Archip %in% c("Palearctic", "Indo-Malay")) %>%
  # number of extinct birds per archipelago
  dplyr::count(Archip) %>% 
  # colonization dates for continents
  dplyr::left_join(colz, by = "Archip") %>%
  # replicate dataframe
  dplyr::slice(rep(1:n(), each = n_samp)) %>% 
  # add run identifier
  dplyr::mutate(run = rep(1:n_samp, n()/n_samp)) %>% 
  # human first arrival from uniform distribution of dates - archipelago level
  dplyr::mutate(hum_unif = purrr::pmap(list(x = hum_upp, y = hum_low), ~ runif(1, .x, .y))) %>% 
  tidyr::unnest(cols = c(hum_unif)) 

cont_pre_excl_spid <- bird %>% 
  # continental fossil extinctions
  dplyr::filter(grp == "cont") %>% 
  # excluding palearctic and indo-malay
  dplyr::filter(!Archip %in% c("Palearctic", "Indo-Malay")) %>%
  dplyr::group_by(Archip) %>% 
  dplyr::mutate(sp_id = 1:n())

cont_pre_excl <- cont_pre_excl_all %>% 
  # create row for every species
  tidyr::uncount(n) %>% 
  # extinction dates from truncated exponential - species level
  dplyr::mutate(pred_date = hum_unif - exp_trunc_cont_split[[2]][1:n()]) %>% 
  # round down to integer, i.e., year
  dplyr::mutate(pred_date = floor(pred_date)) %>% 
  # distribution
  dplyr::mutate(dist = "pre_kno") %>% 
  # add species names
  dplyr::group_by(Archip, run) %>% 
  dplyr::mutate(sp_id = 1:n()) %>% 
  dplyr::left_join(cont_pre_excl_spid, by = c("Archip", "sp_id")) %>% 
  dplyr::select(Archip, run, hum_unif, pred_date, dist, species, commonName)

# palearctic and indo-malay
# extinction date between maximum age of fossil and 1500 CE from uniform distribution - species level
cont_pre_pal <- bird %>% 
  # continental fossil extinctions
  dplyr::filter(grp == "cont") %>% 
  # only palearctic and indo-malay
  dplyr::filter(Archip %in% c("Palearctic", "Indo-Malay")) %>% 
  # colonization dates for continents
  dplyr::left_join(colz, by = "Archip") %>%
  # maximum date of fossil or colonization date
  dplyr::mutate(early_date = purrr::pmap(list(x = fossil_max, y = hum_low), ~ min(.x, .y, na.rm = TRUE))) %>% 
  tidyr::unnest(cols = c(early_date)) %>% 
  # replicate dataframe
  dplyr::slice(rep(1:n(), each = n_samp)) %>% 
  # add run identifier
  dplyr::mutate(run = rep(1:n_samp, n()/n_samp)) %>% 
  # extinction date between maximum age of fossil and 1500 CE from uniform distribution - species level
  dplyr::mutate(hum_unif = purrr::pmap(list(x = hum_upp, y = early_date), ~ runif(1, .x, .y))) %>% 
  tidyr::unnest(cols = c(hum_unif)) %>%
  # round down to integer, i.e., year
  dplyr::mutate(pred_date = floor(hum_unif)) %>% 
  # distribution
  dplyr::mutate(dist = "pre_kno") %>% 
  # select useful columns
  dplyr::select(Archip, run, pred_date, dist)

# combine palearctic with other regions
cont_pre_comb <- dplyr::bind_rows(cont_pre_excl, cont_pre_pal)

# combine all data
ex_da_all <- ex_da %>% 
  # add observed extinct with extinction dates - 182 spp
  dplyr::bind_rows(dates_hist_ext_long) %>% 
  # add observed extinct only recorded from fossils - 17 spp
  dplyr::bind_rows(hist_ins) %>% 
  # add possibly extinct species - ~26 spp
  dplyr::bind_rows(hist_pex_long) %>% 
  # add fossil continental extinctions - 98 spp
  dplyr::bind_rows(cont_pre_comb) %>% 
  # BCE/CE
  dplyr::mutate(pred_date_ce = 1950 - pred_date) %>% 
  # select useful columns
  dplyr::select(Archip, run, pred_date, pred_date_ce, dist, species, commonName)

# # load: ex_da_all
# ex_da_all <- readRDS("data/ex_da_all.rds") # this is the full dataset from the study

# extinction rate through time
ex_da_years_all <- ex_da_all %>%
  group_by(run) %>%
  # number of extinctions per year - n
  dplyr::count(pred_date) %>%
  # add in all years in sequence with zero extinctions
  tidyr::complete(pred_date = tidyr::full_seq(c(-69, 126000), period = 1), fill = list(n = 0)) %>%
  # order by year
  dplyr::arrange(-pred_date) %>%
  # cumulative extinctions per year
  dplyr::mutate(cumsum = cumsum(n)) %>%
  # total extinctions per run
  dplyr::group_by(run) %>%
  dplyr::mutate(tex = sum(n)) %>%
  # number alive per year - extant + all extinct - extinct at time T
  dplyr::mutate(alive = (10865 + tex) - cumsum) %>%
  # extinction rate per year
  dplyr::mutate(ex_rate = n / alive) %>%
  # moving average of 100 years
  dplyr::mutate(ma = zoo::rollmean(ex_rate, k = 100, fill = NA)) %>%
  # replace tiny negative numbers
  dplyr::mutate(ma = ifelse(ma < 0, 0, ma))

ex_da_years_sum <- ex_da_years_all %>%
  # group by year
  group_by(pred_date) %>%
  # calculate median extinction rate across runs per year
  dplyr::summarise(ma_median = median(ma),
                   ma_low = HDInterval::hdi(ma, credMass = 0.95)[[1]],
                   ma_upp = HDInterval::hdi(ma, credMass = 0.95)[[2]]) %>%
  # BCE/CE
  dplyr::mutate(pred_date_ce = 1950 - pred_date)

ex_da_years_cum <- ex_da_years_all %>%
  # group by year
  group_by(pred_date) %>%
  # calculate median extinction rate per year
  dplyr::summarise(cum_median = median(cumsum),
                   cum_low = HDInterval::hdi(cumsum, credMass = 0.95)[[1]],
                   cum_upp = HDInterval::hdi(cumsum, credMass = 0.95)[[2]]) %>%
  # BCE/CE
  dplyr::mutate(pred_date_ce = 1950 - pred_date)

# estimated species extinction dates
sp_date <- ex_da_all %>% 
  dplyr::group_by(species, commonName, Archip) %>% 
  dplyr::summarise_at(vars(pred_date), list(~median(.), ~sd(.))) %>% 
  dplyr::filter(!is.na(species)) %>% 
  dplyr::mutate(median_ce = 1950 - median)

```

## Totals

```{r}

# total

ex_tot <- ex_da_all %>%
  dplyr::ungroup() %>% 
  dplyr::count(run)

ex_tot_hdi <- HDInterval::hdi(ex_tot$n, credMass = 0.95) %>% 
  t() %>% 
  as.data.frame() %>% 
  dplyr::mutate(median = median(ex_tot$n))

ex_tot_hdi

# observed
hist_all <- ex_da_all %>% 
  dplyr::filter(dist == "hist_kno" | dist == "hist_un") %>% 
  dplyr::ungroup() %>% 
  dplyr::count(run) %>% 
  dplyr::mutate(grp = "Observed")

hist_cross_all <- HDInterval::hdi(hist_all$n, credMass = 0.95) %>% 
  t() %>% 
  as.data.frame() %>% 
  dplyr::mutate(median = median(hist_all$n)) %>% 
  dplyr::mutate(grp = "Observed")

hist_cross_all

# fossil
pre_kno_all <- ex_da_all %>% 
  dplyr::filter(dist == "pre_kno") %>% 
  dplyr::ungroup() %>% 
  dplyr::count(run) %>% 
  dplyr::mutate(grp = "Fossil")

pre_kno_cross_all <- HDInterval::hdi(pre_kno_all$n, credMass = 0.95) %>%
  t() %>% 
  as.data.frame() %>% 
  dplyr::mutate(median = median(pre_kno_all$n)) %>% 
  dplyr::mutate(grp = "Fossil")

pre_kno_cross_all

# undiscovered
pre_un_all <- ex_da_all %>% 
  dplyr::filter(dist == "pre_un") %>% 
  dplyr::ungroup() %>% 
  dplyr::count(run) %>% 
  dplyr::mutate(grp = "Undiscovered")

pre_un_cross_all <- HDInterval::hdi(pre_un_all$n, credMass = 0.95) %>% 
  t() %>% 
  as.data.frame() %>% 
  dplyr::mutate(median = median(pre_un_all$n)) %>% 
  dplyr::mutate(grp = "Undiscovered")

pre_un_cross_all

## overall stats ##

# percent extinct
(ex_tot_hdi$median / (10865 + ex_tot_hdi$median)) * 100

# maximum extinction rate
ex_da_years_sum[which.max(ex_da_years_sum$ma_mean),]

# percent unrecorded
(pre_un_cross_all$median / ex_tot_hdi$median) * 100

```

## Main Figures

```{r}

# Map prep

# recenter
center <- 145

# world map centered on 145
mapWorld <- ggplot2::map_data("world", wrap = c(center - 180, center + 180), ylim = c(-60, 85), interior = FALSE)

# recenter islands
isl_pnts <- isl_orig
isl_pnts$long_recenter <- ifelse(isl_pnts$Long < center - 180, isl_pnts$Long + 360, isl_pnts$Long)

# group islands into archipelagos
arch_pnts <- isl_pnts %>%
  dplyr::group_by(Archip) %>%
  # weighted average lat and long - weighted by island area
  dplyr::summarise(mean_long_re = weighted.mean(long_recenter, Area), mean_lat = weighted.mean(Lat, Area))

#### Figure 2 ####

## 2a

# plot total bird extinction violin
swarm_tot <- ggplot2::ggplot(ex_tot, ggplot2::aes(x = 1, y = n)) +
  # half violin
  gghalves::geom_half_violin(side = "r", fill = "grey", colour = "grey") +
  # 95% credible interval
  ggplot2::geom_linerange(ggplot2::aes(ymin = ex_tot_hdi$lower, ymax = ex_tot_hdi$upper), colour = "grey40", lwd = 0.8) +
  # dot for median
  ggplot2::geom_point(ggplot2::aes(y = ex_tot_hdi$median), size = 4, colour = "grey40") +
  # label for median
  ggplot2::annotate("text", x = 1.12, y = ex_tot_hdi$median, label = round(ex_tot_hdi$median, digits = 0), colour = "grey40") +
  # scale across plots
  ggplot2::scale_y_continuous(limits = c(0, (1.05 * max(ex_tot$n))), expand = c(0, 0)) +
  ggplot2::scale_x_continuous(breaks = 1, labels = "Total") +
  # y axis label
  ggplot2::labs(y = "Extinct birds") +
  ggplot2::theme(axis.title.x = ggplot2::element_blank())

# combine summaries
all_bar <- dplyr::bind_rows(hist_cross_all, pre_kno_cross_all, pre_un_cross_all) %>%
  # remove lower and upper values for fossils
  dplyr::mutate_at(vars(c("lower", "upper")), ~ifelse(grp == "Fossil", NA, .)) %>% 
  # order factor for exticntion types
  dplyr::mutate(grp = factor(grp, levels = c("Observed", "Fossil", "Undiscovered")))

# create 2a plot
swarm_dot <- ggplot2::ggplot(all_bar, ggplot2::aes(x = grp, y = median, colour = grp)) +
  # keep plot dimensions
  ggplot2::geom_point(colour = NA) +
  # half violins
  gghalves::geom_half_violin(data = pre_un_all, ggplot2::aes(x = 3, y = n), side = "r", fill = "#ffac6a", colour = "#ffac6a", inherit.aes = FALSE) +
  # 95% credibe intervals
  ggplot2::geom_linerange(ggplot2::aes(ymin = lower, ymax = upper), lwd = 0.8) +
  # dots for medians
  ggplot2::geom_point(size = 4) +
  # labels for medians
  ggplot2::annotate("text", x = 1.15, y = all_bar$median[[1]], label = round(all_bar$median[[1]], digits = 0), colour = "#CC79A7") +
  ggplot2::annotate("text", x = 2.15, y = all_bar$median[[2]], label = round(all_bar$median[[2]], digits = 0), colour = "#0072B2") +
  ggplot2::annotate("text", x = 3.15, y = all_bar$median[[3]], label = round(all_bar$median[[3]], digits = 0), colour = "#D55E00") +
  # scale across plots
  ggplot2::scale_y_continuous(limits = c(0, (1.05 * max(ex_tot$n))), expand = c(0, 0)) +
  # colours
  ggplot2::scale_colour_manual(values = c("#CC79A7", "#0072B2", "#D55E00")) +
  # remove chart junk
  ggplot2::theme(legend.position = "none",
                 axis.title.x = ggplot2::element_blank(),
                 axis.line.y = ggplot2::element_blank(),
                 axis.text.y = ggplot2::element_blank(),
                 axis.title.y = ggplot2::element_blank(),
                 axis.ticks.y = ggplot2::element_blank())

# combine total and dot plot
swarm_tot_dot <- cowplot::plot_grid(swarm_tot, swarm_dot, rel_widths = c(0.5, 1))

## 2b

# map

# observed
hist_all_pie <- ex_da_all %>%   
  # filter to observed extinctions
  dplyr::filter(dist == "hist_kno" | dist == "hist_un") %>% 
  # count extinctions per archipelago per run
  dplyr::group_by(Archip) %>%
  dplyr::count(run) %>% 
  dplyr::group_by(Archip) %>% 
  # median across runs
  dplyr::summarise_at(vars(n), median) %>% 
  dplyr::rename(hist = n)
                      
# fossil
pre_kno_all_pie <- ex_da_all %>% 
  dplyr::filter(dist == "pre_kno") %>% 
  dplyr::group_by(Archip) %>%
  dplyr::count(run) %>% 
  dplyr::group_by(Archip) %>% 
  dplyr::summarise_at(vars(n), median) %>% 
  dplyr::rename(pre_kno = n)

# undiscovered
pre_un_all_pie <- ex_da_all %>%
  dplyr::filter(dist == "pre_un") %>% 
  dplyr::group_by(Archip) %>%
  dplyr::count(run) %>% 
  dplyr::group_by(Archip) %>% 
  dplyr::summarise_at(vars(n), median) %>% 
  dplyr::rename(pre_un = n)

# combine data
all_pie <- dplyr::full_join(hist_all_pie, pre_kno_all_pie, by = "Archip") %>% 
  dplyr::full_join(pre_un_all_pie, by = "Archip")

arch_pnts_swarm <- arch_pnts %>% 
  # join extinction counts to archipelago data
  dplyr::left_join(all_pie, by = "Archip") %>% 
  # fill out zeros
  tidyr::replace_na(list(hist = 0, pre_kno = 0, pre_un = 0)) %>% 
  # total
  dplyr::mutate(tot = hist + pre_kno + pre_un) %>% 
  # total adjusted to area of circle
  dplyr::mutate(tot_a = sqrt(tot/pi))

# create 2b plot
swarm_map <- ggplot2::ggplot(ggplot2::aes(x = long, y = lat), data = mapWorld) +
  # plot polygons for basemap
  ggplot2::geom_polygon(ggplot2::aes(group = group), fill = "grey65") +
  # add pies
  scatterpie::geom_scatterpie(data = arch_pnts_swarm, ggplot2::aes(x = mean_long_re, y = mean_lat, group = Archip, r = tot_a * 1.6), cols = colnames(arch_pnts_swarm)[4:6], alpha = 0.8, colour = NA) +  
  # equal coordinates for map
  ggplot2::coord_equal() +
  # add longitude scale
  ggplot2::scale_x_continuous(expand = c(0, 0), breaks = c(0, 90, 180, 270), labels = c(paste0("0", "째"), paste0("90", "째E"), paste0("180", "째"), paste0("90", "째W"))) +
  # clip latitude
  ggplot2::scale_y_continuous(limits = c(-60, 85)) +
  # colours
  ggplot2::scale_fill_manual(labels = c("Observed ", "Fossil", "Undiscovered"), values = c("#CC79A7", "#0072B2", "#D55E00")) +
  # axis labels
  ggplot2::labs(x = "Longitude", y = "Latitude", size = "Undiscovered prehistoric extinct birds") +
  # legend for pies
  scatterpie::geom_scatterpie_legend((dplyr::filter(arch_pnts_swarm, tot > 0)$tot_a * 1.6), n = 3, x = 215, y = -48, labeller = function(x) signif(pi * (x / 1.6) ^ 2, 1)) +
  # shift legend to bottom and remove legend title
  ggplot2::theme(legend.position = "bottom",
                 legend.title = ggplot2::element_blank())

# combine total, bar and map
swarm_pie <- cowplot::plot_grid(swarm_tot_dot, swarm_map, nrow = 2, ncol = 1, labels = "auto") +
  ggplot2::theme(plot.background = ggplot2::element_rect(fill = "white", colour = "white"))

cowplot::save_plot("figures/fig_2_swarmpie.png", swarm_pie, base_width = 10, base_height = 10, dpi = 300)

#### Figure 3 ####

## 3a

# text size scaler
size_gt <- 6

# cumulative extinctions inset
ski_zoom <- ggplot2::ggplot(ex_da_years_cum, ggplot2::aes(x = pred_date_ce, y = cum_median)) +
  # ribbon of uncertainty
  ggplot2::geom_ribbon(ggplot2::aes(ymin = cum_low, ymax = cum_upp), fill = "lightgrey") +
  # median line
  ggplot2::geom_path(lwd = 1.5, show.legend = FALSE) +
  # clip plot and specify tidy breaks
  ggplot2::scale_x_continuous(limits = c(-5000, (max(ex_da_years_sum$pred_date_ce) * 1.1)), breaks = c(-4000, -2000, 0, 2000), expand = c(0, 0), name = "Year") +
  # specify tidy breaks with commas
  ggplot2::scale_y_continuous(expand = c(0, 0), breaks = c(0, 200, 400, 600, 800, 1000, 1200, 1400, 1600, 1800), labels = scales::comma, name = "Cumulative bird extinctions")

# extinction rate through time
wiggle_zoom <- ggplot2::ggplot(ex_da_years_sum, ggplot2::aes(x = pred_date_ce, y = ma_median)) +
  # baseline extinctions dashed line
  ggplot2::geom_hline(yintercept = 0.2e-05, lty = 2) +
  # ribbon of uncertainty
  ggplot2::geom_ribbon(ggplot2::aes(ymin = ma_low, ymax = ma_upp), fill = "lightgrey") +
  # median line
  ggplot2::geom_path(lwd = 1.5, show.legend = FALSE) +
  # add lines and text to show position of EX1
  ggplot2::geom_segment(ggplot2::aes(x = -1541, y = (4.395691e-05 + 4e-06), xend = 438, yend = (4.395691e-05 + 4e-06)), lwd = 1.1, colour = "#CC79A7") +
  ggplot2::geom_segment(ggplot2::aes(x = -1541, y = (4.395691e-05 + 5e-06), xend = -1541, yend = (4.395691e-05 + 3e-06)), lwd = 1.1, colour = "#CC79A7") +
  ggplot2::geom_segment(ggplot2::aes(x = 438, y = (4.395691e-05 + 5e-06), xend = 438, yend = (4.395691e-05 + 3e-06)), lwd = 1.1, colour = "#CC79A7") +
  ggplot2::geom_text(ggplot2::aes(x = -551.5, y = (4.395691e-05 + 8e-06), label = "EX1"), colour = "#CC79A7", size = size_gt) +
  # add lines and text to show position of EX2
  ggplot2::geom_segment(ggplot2::aes(x = 459, y = (0.0001566098 + 4e-06), xend = 1742, yend = (0.0001566098 + 4e-06)), lwd = 1.1, colour = "#D55E00") +
  ggplot2::geom_segment(ggplot2::aes(x = 459, y = (0.0001566098 + 5e-06), xend = 459, yend = (0.0001566098 + 3e-06)), lwd = 1.1, colour = "#D55E00") +
  ggplot2::geom_segment(ggplot2::aes(x = 1742, y = (0.0001566098 + 5e-06), xend = 1742, yend = (0.0001566098 + 3e-06)), lwd = 1.1, colour = "#D55E00") +
  ggplot2::geom_text(ggplot2::aes(x = 1100.5, y = (0.0001566098 + 8e-06), label = "EX2"), colour = "#D55E00", size = size_gt) +
  # add lines and text to show position of EX3
  ggplot2::geom_segment(ggplot2::aes(x = 1742, y = (8.246472e-05 + 4e-06), xend = 2019, yend = (8.246472e-05 + 4e-06)), lwd = 1.1, colour = "#0072B2") +
  ggplot2::geom_segment(ggplot2::aes(x = 1742, y = (8.246472e-05 + 5e-06), xend = 1742, yend = (8.246472e-05 + 3e-06)), lwd = 1.1, colour = "#0072B2") +
  ggplot2::geom_segment(ggplot2::aes(x = 2019, y = (8.246472e-05 + 5e-06), xend = 2019, yend = (8.246472e-05 + 3e-06)), lwd = 1.1, colour = "#0072B2") +
  ggplot2::geom_text(ggplot2::aes(x = 1881, y = (8.246472e-05 + 8e-06), label = "EX3"), colour = "#0072B2", size = size_gt) +
  # clip plot and specify tidy breaks
  ggplot2::scale_x_continuous(limits = c(-5000, (max(ex_da_years_sum$pred_date_ce) * 1.1)), breaks = c(-5000, -4000, -3000, -2000, -1000, 0, 1000, 2000), expand = c(0, 0), name = "Year") +
  # fancy scientific labels
  ggplot2::scale_y_continuous(limits = c(0, (max(ex_da_years_sum$ma_upp, na.rm = TRUE) * 1.1)), expand = c(0, 0), labels = fancy_scientific, name = "Bird extinction rate (extinctions/species/year)") +
  # tidy plot
  ggplot2::theme(plot.title = ggplot2::element_text(hjust = 0.5),
                 text = ggplot2::element_text(size = 20),
                 axis.text = ggplot2::element_text(size = 15))

# combine extinction plots
wig_c <- cowplot::ggdraw() +
  # extinction rate through time
  cowplot::draw_plot(wiggle_zoom) +
  # inset cumulative extinctions
  cowplot::draw_plot(ski_zoom, x = 0.2, y = 0.4, width = 0.4, height = 0.4)

# use this code (and adapt numbers) to find peaks and boundaries of extinction events
test <- dplyr::filter(ex_da_years_sum, pred_date_ce > 2000 & pred_date_ce < 0)

## 3b
# extinction events maps

# extinction events are based on full dataset not simplified dataset

# count extinctions per archipelago for extinction event 1
b1 <- ex_da_all %>%
  # filter to years of extinction event
  dplyr::filter(pred_date_ce > -1541 & pred_date_ce < 438) %>%
  # count extinctions and sort archipelagos
  dplyr::count(Archip, sort = TRUE) %>%
  # convert into actual numbers of extinctions
  dplyr::mutate(count = n / n_samp) 

# count extinctions per archipelago for extinction event 2
b2 <- ex_da_all %>%
  dplyr::filter(pred_date_ce > 459 & pred_date_ce < 1742) %>%
  dplyr::count(Archip, sort = TRUE) %>%
  dplyr::mutate(count = n / n_samp) 

# count extinctions per archipelago for extinction event 3
b3 <- ex_da_all %>%
  dplyr::filter(pred_date_ce > 1742) %>%
  dplyr::count(Archip, sort = TRUE) %>%
  dplyr::mutate(count = n / n_samp)

# combine
b_all <- rbind(b1, b2, b3)

# rank archipelagos by extinctions within extinction event 1
ap_b1 <- arch_pnts %>%
  # join extinction counts
  dplyr::right_join(b1, by = "Archip") %>%
  # rank archipelagos by number of extinctions
  dplyr::mutate(rank = rank(desc(count), ties.method = "first")) %>%
  # get names of top 5 archipelagos
  dplyr::mutate(Archip_neat = ifelse(rank < 6, Archip, "")) %>%
  # get rank numbers 1 to 5
  dplyr::mutate(rank = ifelse(rank < 6, rank, "")) %>% 
  dplyr::arrange(rank) %>% 
  dplyr::mutate(Archip_neat = factor(Archip_neat, levels = unique(.$Archip_neat)))

# rank archipelagos by extinctions within extinction event 2
ap_b2 <- arch_pnts %>%
  dplyr::right_join(b2, by = "Archip") %>%
  dplyr::mutate(rank = rank(desc(count), ties.method = "first")) %>%
  dplyr::mutate(Archip_neat = ifelse(rank < 6, Archip, "")) %>%
  dplyr::mutate(rank = ifelse(rank < 6, rank, "")) %>% 
  dplyr::arrange(rank) %>% 
  # convert to factor
  dplyr::mutate(Archip_neat = factor(Archip_neat, levels = unique(.$Archip_neat)))

# rank archipelagos by extinctions within extinction event 3
ap_b3 <- arch_pnts %>%
  dplyr::right_join(b3, by = "Archip") %>%
  dplyr::mutate(rank = rank(desc(count), ties.method = "first")) %>%
  dplyr::mutate(Archip_neat = ifelse(rank < 6, Archip, "")) %>%
  dplyr::mutate(rank = ifelse(rank < 6, rank, "")) %>% 
  dplyr::arrange(rank) %>% 
  dplyr::mutate(Archip_neat = factor(Archip_neat, levels = unique(.$Archip_neat)))

# scaler for range of point sizes
scaler <- c(3, 20)
# transparency
alpha <- 0.5

b1_map <- ggplot2::ggplot(ggplot2::aes(x = long, y = lat), data = mapWorld) +
  # polygons for basemap
  ggplot2::geom_polygon(ggplot2::aes(group = group), fill = "grey65") +
  # points showing extinctions
  ggplot2::geom_point(data = ap_b1, ggplot2::aes(x = mean_long_re, y = mean_lat, size = count), colour = "#CC79A7", alpha = alpha, show.legend = FALSE) +
  # label ranks
  ggplot2::geom_text(data = ap_b1, ggplot2::aes(x = mean_long_re, y = mean_lat, label = rank), size = 4) +
  # title and y axis title
  ggplot2::labs(title = "1541 BCE - 438 CE", y = "EX1", x = "") +
  ggplot2::scale_discrete_identity(aesthetics = "label", name = "", breaks = as.factor(ap_b1$rank), labels = ap_b1$Archip_neat, guide = "legend") +
  # clip latitude
  ggplot2::scale_y_continuous(limits = c(-60, 85)) +
  # scale size of points
  ggplot2::scale_size(range = scaler, limits = c(min(b_all$count), max(b_all$count))) +
  # equal coordinates for map
  ggplot2::coord_equal() +
  # tidy plot
  ggplot2::theme(legend.title = ggplot2::element_blank(),
                 axis.title.y = ggplot2::element_text(colour = "#CC79A7", angle = 0, vjust = 0.5),
                 axis.text = ggplot2::element_blank(),
                 axis.ticks = ggplot2::element_blank(),
                 axis.line = ggplot2::element_blank(),
                 plot.title = ggplot2::element_text(hjust = 0.5),
                 axis.title = ggplot2::element_text(size = 20),
                 legend.justification = "top")

b2_map <- ggplot2::ggplot(ggplot2::aes(x = long, y = lat), data = mapWorld) +
  ggplot2::geom_polygon(ggplot2::aes(group = group), fill = "grey65") +
  ggplot2::geom_point(data = ap_b2, ggplot2::aes(x = mean_long_re, y = mean_lat, size = count), colour = "#D55E00", alpha = alpha, show.legend = FALSE) +
  ggplot2::geom_text(data = ap_b2, ggplot2::aes(x = mean_long_re, y = mean_lat, label = rank), size = 4) +
  ggplot2::labs(title = "459 CE - 1742 CE", y = "EX2", x = "") +
  ggplot2::scale_discrete_identity(aesthetics = "label", name = "", breaks = as.factor(ap_b2$rank), labels = ap_b2$Archip_neat, guide = "legend") +
  ggplot2::scale_y_continuous(limits = c(-60, 85)) +
  ggplot2::scale_size(range = scaler, limits = c(min(b_all$count), max(b_all$count))) +
  ggplot2::coord_equal()  +
  ggplot2::theme(legend.title = ggplot2::element_blank(),
                 axis.title.y = ggplot2::element_text(colour = "#D55E00", angle = 0, vjust = 0.5),
                 axis.text = ggplot2::element_blank(),
                 axis.ticks = ggplot2::element_blank(),
                 axis.line = ggplot2::element_blank(),
                 plot.title = ggplot2::element_text(hjust = 0.5),
                 axis.title = ggplot2::element_text(size = 20),
                 legend.justification = "top",
                 legend.margin = ggplot2::margin(c(5,5,5,1)))

b3_map <- ggplot2::ggplot(ggplot2::aes(x = long, y = lat), data = mapWorld) +
  ggplot2::geom_polygon(ggplot2::aes(group = group), fill = "grey65") +
  ggplot2::geom_point(data = ap_b3, ggplot2::aes(x = mean_long_re, y = mean_lat, size = count), colour = "#0072B2", alpha = alpha, show.legend = FALSE) +
  ggplot2::geom_text(data = ap_b3, ggplot2::aes(x = mean_long_re, y = mean_lat, label = rank), size = 4) +
  ggplot2::labs(title = "1742 CE - 2019 CE", y = "EX3", x = "") +
  ggplot2::scale_discrete_identity(aesthetics = "label", name = "", breaks = as.factor(ap_b3$rank), labels = ap_b3$Archip_neat, guide = "legend") +
  ggplot2::scale_y_continuous(limits = c(-60, 85)) +
  ggplot2::scale_size(range = scaler, limits = c(min(b_all$count), max(b_all$count))) +
  ggplot2::coord_equal()  +
  ggplot2::theme(legend.title = ggplot2::element_blank(),
                 axis.title.y = ggplot2::element_text(colour = "#0072B2", angle = 0, vjust = 0.5),
                 axis.text = ggplot2::element_blank(),
                 axis.ticks = ggplot2::element_blank(),
                 axis.line = ggplot2::element_blank(),
                 plot.title = ggplot2::element_text(hjust = 0.5),
                 axis.title = ggplot2::element_text(size = 20),
                 legend.justification = "top")

# legend for maps
b_leg <- ggplot2::ggplot(ggplot2::aes(x = long, y = lat), data = mapWorld) +
  ggplot2::geom_point(data = ap_b1, ggplot2::aes(x = mean_long_re, y = mean_lat, size = count), colour = "grey", alpha = alpha, show.legend = TRUE) +
  ggplot2::labs(size = "Extinctions") +
  ggplot2::scale_y_continuous(limits = c(-60, 85)) +
  # choose breaks for legedn of point sizes
  ggplot2::scale_size(range = scaler, limits = c(min(b_all$count), max(b_all$count)), breaks = c(5, 30, 100)) +
  ggplot2::coord_equal() +
  ggplot2::theme(legend.direction = "horizontal") +
  ggplot2::guides(size = ggplot2::guide_legend(title.position = "top", title.hjust = 0.5))

# get legend for maps
b_leg <- cowplot::get_legend(b_leg)

# combine extinction event maps
ee <- cowplot::plot_grid(b1_map, b2_map, b3_map, nrow = 3, labels = c("b", "c", "d"))

# combine extinction event maps and legend
ee_leg <- cowplot::ggdraw() +
  # extinction event maps
  cowplot::draw_plot(ee) +
  # map legend
  cowplot::draw_plot(b_leg, x = 0.71, y = -0.1, width = 0.3, height = 0.3)
  
# combine exticntion rate plot and extiontion event maps plot
wig_window <- cowplot::plot_grid(wig_c, ee_leg, nrow = 1, ncol = 2, labels = c("a", "")) +
  ggplot2::theme(plot.background = ggplot2::element_rect(fill = "white", colour = "white"))

cowplot::save_plot("figures/fig_3_wiggle.png", wig_window, base_width = 18, base_height = 10, dpi = 300, bg = "white")

```

## Lower bound model

```{r}

# species description dates
dates_found_orig <- readr::read_csv("data/dates_found.csv")
# species = Scientific name; ref.date = Reference for date described; authority = Taxonomic authority; described = Year described

# description dates for lower bound - cumulative description rate
dates_descr <- dplyr::count(dates_found_orig, described) %>%
  dplyr::mutate(cumsum = cumsum(n))

# functions for modelling logisitic - Daniele Silvestro
get_mu0_logistic <- function(L, k, x0, x){
	return(L / (1 + exp(-k * (x - x0))))
}

sliding_window <- function(i, wsize = 0.5){
	new_i <- i + (runif(1) - 0.5) * wsize
	return(new_i)
}

multiplier_proposal <- function(i, d = 1.1){
	u <- runif(1)
	l <- 2 * log(d)
	m <- exp(l * (u - 0.5))
 	ii <- i * m
	hastings_ratio <- log(m)
	return(c(ii, hastings_ratio))
}

# time steps
x <- dates_descr$described - min(dates_descr$described)

# cumulative description
dat <- dates_descr$cumsum

maxDat <- max(dat)
# init parameters
epsilonA <- 2   # error (variance)
LA <- maxDat * 10 # maximum of the logistic = 'dy + L'
dyA <- 0        # shifts up or down the logistic
kA <- 2         # slope
x0A <- min(x)   # mid point

mu0 <- get_mu0_logistic(LA, kA, x0A, x) + dyA

likA <- sum(dnorm(dat, mu0, epsilonA, log = TRUE))

res <- NULL

# 1,000,000 iterations for example
for (it in 0:1000000){
	L  <- LA
	k  <- kA
	x0 <- x0A
	dy <- dyA
	epsilon <- epsilonA
	hasting <- 0
	r <- runif(5)
	if (r[1] < 0.1) {
		temp = multiplier_proposal(LA, 1.2)
		L = temp[1]
		hasting = hasting + temp[2]
	}
	if (r[2] < 0.1) {
		temp = multiplier_proposal(kA)
		k = temp[1]
		hasting = hasting + temp[2]
	}
	if (r[3] < 0.1) {
		x0 = sliding_window(x0A, 5)
		if (x0 > max(x)){
			x0 = max(x)- (x0 - max(x))
		}
	}
	if (r[4] < 0.1) {
		dy = sliding_window(dyA, 2)
	}
	if (r[5] < 0.1) {
		temp = multiplier_proposal(epsilon, 1.2)
		epsilon = temp[1]
		hasting = hasting + temp[2]
	}
	mu0 = get_mu0_logistic(L, k, x0, x) + dy
	lik = sum(dnorm(dat, mu0, epsilon, log = TRUE))
	if (it %% 5000 == 0){
		res = rbind(res, c(it, likA, LA, kA, x0A, dyA, epsilonA))
	}
	if ((lik - likA + hasting) > log(runif(1))){
		LA  = L
		kA  = k
		x0A = x0
		dyA = dy
		epsilonA = epsilon
		likA = lik
	}
}

# tidy output data
res <- as.data.frame(res) %>% 
  setNames(c("it", "likA", "LA", "kA", "x0A", "dyA", "epsilonA"))

# # simple plot of estimated results
# plot(dates_descr$described, dat); lines(dates_descr$described, get_mu0_logistic(LA, kA, x0A, x) + dyA)

# sample from lower bound

# lower bound estimates
# posterior distribution
low_samp <- res[11:nrow(res),] %>%
  # subtract described species from describable species to get undescribed species
  dplyr::mutate(low_spp = LA - sum(dates_descr$n)) %>%
  dplyr::mutate(dist = "low")

# # load: low_samp
# low_samp <- readRDS("data/low_samp_1000.rds") # this is the full dataset from the study

df <- cbind(dates_descr[,1], dat, mu0) %>% 
  as.data.frame()

p1 <- ggplot2::ggplot(df, ggplot2::aes(x = described, y = dat)) +
  ggplot2::geom_point(colour = "black") +
  ggplot2::geom_line(ggplot2::aes(x = described, y = mu0), colour = "black") +
  ggplot2::scale_y_continuous(limits = c(0, max(low_samp$LA) + 20), breaks = scales::pretty_breaks(n = 5)) +
  ggplot2::labs(x = "Year described", y = "Fossil extinct birds")

p2 <- ggplot2::ggplot(low_samp, ggplot2::aes(x = dist, y = LA)) +
  gghalves::geom_half_violin(side = "r", fill = "#009E73", colour = "#009E73") +
  ggplot2::geom_point(data = data.frame(dist = "low", LA = median(low_samp$LA)), colour = "#009E73", size = 5, shape = 15) +
  ggplot2::ylim(0, max(low_samp$LA) + 20) +
  ggplot2::scale_x_discrete(labels = c("Describable species")) +
  ggplot2::theme(panel.border = ggplot2::element_blank(), 
                 axis.line = ggplot2::element_line(), 
                 axis.line.y = ggplot2::element_blank(), 
                 axis.title = ggplot2::element_blank(), 
                 axis.text.y = ggplot2::element_blank(), 
                 axis.ticks.y = ggplot2::element_blank(), 
                 legend.position = "none")

p_comb <- cowplot::plot_grid(p1, p2, rel_widths = c(1, 0.6), align = "h") +
  ggplot2::theme(plot.background = ggplot2::element_rect(fill = "white", colour = "white"))

cowplot::save_plot("figures/fig_5_lowspp.png", p_comb, base_width = 12, base_height = 7, dpi = 300)

median(low_samp$low_spp)

```

